Query Optimization for Dynamic Imputation

### 2017-05-10 paper/algorithm.tex

\subsection{Imputation operators}
\label{sec:operators}
We introduce two new relational operators to perform imputation: \textit{Impute} ($\mu$) and
\textit{Drop} ($\delta$). Each operator takes arguments $(C, R)$ where $C$ is a set of
attributes and $R$ is a relation. \textit{Impute} uses a machine learning algorithm to
replace all \nullv{} values with non-\nullv{} values for attributes $C$ in the relation $R$ (discussed in detail in~\Cref{sec:imputation}).
In our experiments, we implement \textit{Impute} using a chained-equation decision trees algorithm, but our system is agnostic to this choice.
\textit{Drop} simply removes from $R$ all tuples which have a \nullv{} value for some attribute in $C$.
Both operators guarantee that the resulting relation will contain no \nullv{} values for
attributes in $C$.

@@

\subsection{Cost model}
\label{sec:cost-model}
The cost of a plan $q$ is a tuple $\langle \textsc{Penalty}(q), \textsc{Time}(q) \rangle$.
$\textsc{Penalty}(q) \in [0,1]$ is an estimate of the amount of information lost by the imputation procedures used.
$\textsc{Time}(q) \in (0, \infty)$ is an estimate of the runtime of a query $q$ derived from table statistics and selectivity estimation of the query predicates.

\subsubsection{Pareto-optimal plans}
Given a query, \ProjectName{} produces a final set of plans $P$ by only keeping plans that are not dominated by others in the search space. So for a search space $S$, \[P= \{ p ~|~ \not\exists p' \in S.\ p \neq p' \land \doms{p'}{p} \}.\]

\begin{definition}
We say that a plan $p$ with cost $\langle l, t \rangle$ \emph{dominates} a plan $p'$ with cost $\langle l', t' \rangle$ if $\textsc{Dirty}(p) = \textsc{Dirty}(p') \land ((l \leq l' \land t < t') \lor (l < l' \land t \leq t'))$. We denote this as $\doms{p}{p'}$. 
\end{definition}

This set $P$ will be the Pareto frontier of $S$~\cite{pareto1964cours}, and it contains the best option for all possible trade-offs of $\textsc{Time}$ and $\textsc{Penalty}$ for the current query.

@@

\begin{definition}
Let $P$ be a set of plans. For $p \in P$, we say $p$ is $\alpha$-bound, if $(\textsc{Penalty}(p) - \min_{p' \in P}(\textsc{Penalty}(p')) \leq \alpha$. We denote this as $p^\alpha$.
We define the set of $\alpha$-bound plans in $P$ as $P^\alpha$.
We say a plan $p^\alpha$ is $\alpha$-bound optimal if $\forall p' \in P^\alpha. \textsc{Time}(p) \leq \textsc{Time}(p')$. 
\end{definition}

Given $\alpha$ and $P$, \ProjectName{} returns an $\alpha$-bound optimal plan in $P$.

@@

\subsubsection{Cardinality estimation}
\label{sec:cardinal}
The computation of $\textsc{Time}(q)$ and $\textsc{Penalty}(q)$ rely accurate cardinality estimates for the plan $q$ and its sub-plans.
These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational algebra, but also by the imputation operators.
For example, a drop operator will reduce the cardinality of the result while an impute operator will maintain the same cardinality as the input.

@@

$\textsc{Penalty}(q, P)$ assigns a cost in $(0,1]$ to each imputation operator in the plan $q$ according to its estimated quality, then averages over the individual costs.
For example, each occurrence of $\delta$ incurs a cost of 1 (i.e.\ we lose all information that dropped tuples might have encoded).
Imputing \nullv{} fields with $\mu$ incurs a cost in $(0, 1]$ according to a cost function $P$, which is specific to the imputation method in use.

The choice of $P$ depends on the properties of the imputation method being used.
A well chosen $P$ should provide a reasonable estimate of the quality of imputation from the histogram describing the input data.
From this histogram, $P$ can obtain the number of attributes in its input, its cardinality, and an estimate of its variance.
Intuitively, imputation accuracy \textit{increases} when more complete attributes and complete tuples are available; correspondingly, the $P$ should \textit{decrease} as the number of values increases.
$P$ should never decrease to zero, as this would mean that it can perform perfect imputation.
It should reflect the intuition that the number of training observations has diminishing returns by decreasing more slowly as the number of values increases.
In Section~\ref{sec:imputation}, we discuss choices for $P$ for several common imputation methods.

>>>>

\subsection{Imputation operators}
\label{sec:operators}
We introduce two new relational operators to perform imputation: \textit{Impute} ($\mu$) and
\textit{Drop} ($\delta$). Each operator takes arguments $(C, R)$ where $C$ is a set of
attributes and $R$ is a relation. \textit{Impute} uses a machine learning algorithm to
replace all \nullv{} values with non-\nullv{} values for attributes $C$ in the relation $R$ (discussed in detail in~\Cref{sec:imputation}).
In our experiments, we implement \textit{Impute} using a general-purpose chained-equation decision trees algorithm, but our system is agnostic to this choice.
\textit{Drop} simply removes from $R$ all tuples which have a \nullv{} value for some attribute in $C$.
Both operators guarantee that the resulting relation will contain no \nullv{} values for
attributes in $C$.

@@

\subsection{Cost model}
\label{sec:cost-model}
The cost of a plan $q$ is a tuple $\langle \textsc{Penalty}(q), \textsc{Time}(q) \rangle$.
$\textsc{Penalty}(q) \in [0,1]$ is a heuristic that attempts to quantify the amount of information lost by the imputation procedures used.
$\textsc{Time}(q) \in (0, \infty)$ is an estimate of the runtime of a query $q$ derived from table statistics and selectivity estimation of the query predicates.

\subsubsection{Pareto-optimal plans}
Given a query, \ProjectName{} produces a final set of query plans $Q$ by only keeping plans that are not dominated by others in the search space. So for a search space $S$, \[Q= \{ q ~|~ \not\exists q' \in S.\ q \neq q' \land \doms{q'}{q} \}.\]

\begin{definition}
We say that a plan $q$ with cost $\langle l, t \rangle$ \emph{dominates} a plan $q'$ with cost $\langle l', t' \rangle$ if $\textsc{Dirty}(q) = \textsc{Dirty}(q') \land ((l \leq l' \land t < t') \lor (l < l' \land t \leq t'))$. We denote this as $\doms{q}{q'}$. 
\end{definition}

This set $Q$ will be the Pareto frontier of $S$~\cite{pareto1964cours}, and it contains the best option for all possible trade-offs of $\textsc{Time}$ and $\textsc{Penalty}$ for the current query.

@@

\begin{definition}
Let $Q$ be a set of plans. For $q \in Q$, we say $q$ is $\alpha$-bound, if $(\textsc{Penalty}(q) - \min_{q' \in Q}(\textsc{Penalty}(q')) \leq \alpha$. We denote this as $q^\alpha$.
We define the set of $\alpha$-bound plans in $Q$ as $Q^\alpha$.
We say a plan $q^\alpha$ is $\alpha$-bound optimal if $\forall q' \in Q^\alpha. \textsc{Time}(q) \leq \textsc{Time}(q')$. 
\end{definition}

Given $\alpha$ and $Q$, \ProjectName{} returns an $\alpha$-bound optimal plan in $Q$.

@@

\subsubsection{Cardinality estimation}
\label{sec:cardinal}
The computation of $\textsc{Penalty}(q)$ and$\textsc{Time}(q)$ relies on accurate cardinality estimates for the plan $q$ and its sub-plans.
These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational algebra, but also by the imputation operators.
For example, a drop operator will reduce the cardinality of the result while an impute operator will maintain the same cardinality as the input.

@@

$\textsc{Penalty}(q, P)$ assigns a penalty in $(0,1]$ to each imputation operator in the
plan $q$ according to its estimated quality, then averages over the individual penalties.
For example, each occurrence of $\delta$ incurs a penalty of 1 (i.e.\ we lose all information that dropped tuples might have encoded).
Imputing \nullv{} fields with $\mu$ incurs a penalty in $(0, 1]$ according to a cost function $P$, which is specific to the imputation method in use.

The choice of $P$ depends on the properties of the imputation method being used.
A well-chosen $P$ should provide a reasonable estimate of the quality of imputation from the histogram describing the input data.
From this histogram, $P$ can obtain the number of attributes in its input, its cardinality, and an estimate of its variance.
Intuitively, imputation accuracy \textit{increases} when more complete attributes and complete tuples are available; correspondingly, the $P$ should \textit{decrease} as the number of values increases.
$P$ should never decrease to zero, as this would mean that it can perform perfect imputation.
It should reflect the intuition that the number of training observations has diminishing returns by decreasing more slowly as the number of values increases.
In Section~\ref{sec:imputation}, we discuss choices for $P$ for several common imputation methods.
